---
title: 'SVbyEye: A visual approach to study genomic structural variantion.'
author: "David Porubsky"
date: "`r Sys.Date()`"
package: SVbyEye
bibliography: SVbyEyeBib.bib
abstract: >
  To be filled in.
vignette: >
  %\\VignetteIndexEntry{Vignette Title} 
  %\\VignetteEngine{knitr::rmarkdown}
  %\\VignetteEncoding{UTF-8}
  %\\newenvironment{CSLReferences}
output:
  BiocStyle::pdf_document
---

# Introduction

SVbyEye relies on assembly to reference alignments produced by tools such as minimap2 [@Li2016-ha] or other tools able to report alignments in a PAF format.


\newpage

# Functionalities
Here are the minimal parameters required to successfully run genome scaffolding using SaaRclust.

\textbf{Low-level functions:} List of function... \hfill \break
\textbf{High-level functions:} List of function... \hfill \break

For more details on available functions and their parameters, please run.
```
library(SVbyEye)
help(package="SVbyEye")
```
For more details on function specific parameters, please run.
```
library(SVbyEye)
?<function_name> (e.g. ?filterPaf)
```

# Quick Start

To demonstrate the functionality of SVbyEye ...

```{r, message=FALSE}
## Load the SVbyEye package
library(SVbyEye)
```

Run the code below in order to create simple plot.
```{r, echo=TRUE, message=FALSE, fig.width=10, fig.height=3, fig.align="left"}
## Get PAF to plot
paf.file <- system.file("extdata", "test1.paf", 
                        package="SVbyEye")
## Read in PAF 
paf.table <- readPaf(paf.file = paf.file, 
                     include.paf.tags = TRUE, restrict.paf.tags = 'cg')
## Make a plot
## Color by alignment directionality
plotMiro(paf.table = paf.table, color.by = 'direction')
```

\newpage

# Session Info
```{r session_info, include=TRUE, echo=TRUE, results='markup'}
devtools::session_info()
```

Report any issues
[here](https://github.com/daewoooo/SVbyEye/issues):

\newpage

# References
